---
title: "Lead solubility prediction using R and Shiny"
description: |
  An R package and Shiny app for PHREEQC-based equilibrium lead and copper solubility prediction.
author:
  - name: Ben Trueman
    url: {}
date: 06-18-2021
bibliography: references.bib
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library("tidyverse")
library("pbcusol")
theme_set(
  theme_bw(14) + 
    theme(
      strip.background = element_blank(),
      panel.grid = element_blank()
    )
)
palette <- wesanderson::wes_palette("Zissou1")
```

Recently, I've been experimenting with the excellent R package `tidyphreeqc` [@tidyphreeqc], which provides a convenient interface for generating PHREEQC input files [@phreeqc_r; @phreeqc]. `tidyphreeqc` incorporates easily into my workflow, and it's found its way into several recent publications [@li_impact_2021; @li_controlling_2020].

I've also used it to build a separate package, [`pbcusol`](https://github.com/bentrueman/pbcusol), that handles lead and copper solubility prediction specifically, with curated thermodynamic data and some domain specific features implemented in what is hopefully a straightforward collection of functions. `pbcusol` is implemented as a Shiny app, available [here](https://bentrueman.shinyapps.io/shinypbcusol/).

`pbcusol` can be used to generate lead and copper solubility predictions that are comparable with those found in literature [@schock1996corr], as detailed in the package README on [GitHub](https://github.com/bentrueman/pbcusol). (They're reproduced here using the `wesanderson` package for colour palettes [@wes].)

```{r basic-grid, fig.height=3, cache=TRUE}

dic_increment_cu <- 1.5
solutions_cu <- list("Tenorite", "Malachite") %>%
  set_names() %>%
  map_dfr(
    ~eq_sol(
      element = "Cu",
      ph = seq(6.5, 10.5, by = .025),
      dic = seq(1, 150, by = dic_increment_cu),
      phase = .x
    ),
    .id = "phase"
  )

dic_increment_pb <- .8
solutions_pb <- list("Cerussite", "Hydcerussite", "Hxypyromorphite") %>%
  set_names() %>%
  map_dfr(
    ~ eq_sol(
      element = "Pb",
      ph = seq(6.5, 10.5, by = .025),
      dic = seq(1, 80, by = dic_increment_pb),
      phosphate = .16,
      phase = .x
    ),
    .id = "phase"
  )

list(
  "[Pb]" = rename(solutions_pb, conc_ppb = pb_ppb), 
  "[Cu]" = rename(solutions_cu, conc_ppb = cu_ppb)
) %>% 
  bind_rows(.id = "element") %>% 
  mutate(
    dic_ppm = if_else(
      element == "[Pb]", 
      plyr::round_any(dic_ppm, dic_increment_pb), 
      plyr::round_any(dic_ppm, dic_increment_cu)
    )
  ) %>% 
  group_by(element, pH, dic_ppm) %>% 
  summarize(
    phase = phase[which.min(conc_ppb)],
    conc_ppb = min(conc_ppb) %>% log10()
  ) %>% 
  ungroup() %>% 
  ggplot(aes(x = dic_ppm, y = pH)) + 
  facet_wrap(vars(element), scales = "free_x") + 
  geom_point(aes(col = conc_ppb), shape = 15, size = 1) +
  geom_contour(
    aes(z = conc_ppb), col = "white", linetype = 2, 
    size = .2, breaks = seq(-4, 3, by = .1)
  ) + 
  geom_contour(
    data = function(x) x %>% 
      filter(element == "[Cu]"),
    aes(z = conc_ppb), col = "white", breaks = -3:3
  ) + 
  geom_smooth(
    data = function(x) x %>% 
      filter(phase == "Tenorite") %>% 
      group_by(element, dic_ppm) %>% 
      summarize(min = min(pH)),
    aes(x = dic_ppm, y = min),
    linetype = 2, size = 1, col = "white", se = FALSE, span = .4
  ) + 
  geom_smooth(
    data = function(x) x %>% 
      filter(phase == "Hydcerussite") %>% 
      group_by(element, dic_ppm) %>% 
      summarize(min = min(pH)),
    aes(x = dic_ppm, y = min),
    linetype = 2, size = 1, col = "white", se = FALSE, span = .4
  ) +
  geom_smooth(
    data = function(x) x %>% 
      filter(phase == "Hxypyromorphite") %>% 
      group_by(element, dic_ppm) %>% 
      summarize(max = max(pH)) %>% 
      filter(dic_ppm > 35),
    aes(x = dic_ppm, y = max),
    linetype = 2, size = 1, col = "white", se = FALSE, span = .4
  ) +
  scale_colour_gradientn(
    colours = palette, 
    labels = function(breaks) 1e-3 * (10 ^ breaks)
  ) +
  labs(
    x = expression("[C]"[inorganic]~"(mg L"^"-1"*")"),
    y = "pH",
    col = expression("[element] (mg L"^"-1"*")")
  ) +
  geom_label(
    data = tibble(
      element = c(rep("[Pb]", 3), rep("[Cu]", 2)),
      label = c(
        "PbCO[3]", 
        "Pb[3]*'('*CO[3]*')'[2]*'('*OH*')'[2]", 
        "Pb[5]*'('*PO[4]*')'[3]*OH", 
        "Cu[2]*'('*CO[3]*')'*'('*OH*')'[2]", 
        "CuO"
      ),
      x = c(65, 40, 25, 100, 50),
      y = c(7.6, 10, 7, 7.5, 10.5)
    ),
    aes(x = x, y = y, label = label),
    parse = TRUE,
    label.r = unit(0, "cm"), label.size = unit(0, "cm")
  )
```

`pbcusol` can also be used to generate predicted lead and copper solubility in the presence of humic substances. This is also outlined in the package README, but here is an example of the type of visualization that can be generated. Of course, these predictions should be properly validated---more on that later.

```{r nom-grid, cache=TRUE}

grid_dim <- 50 # dimension of solubility grid square

future::plan("multisession")

nom_grid <- crossing(
  pH_in = seq(6.5, 10.5, length.out = grid_dim),
  nom_in = seq(0, 3e-3, length.out = grid_dim)
) %>% 
  rowid_to_column() %>% 
  group_by(rowid) %>% 
  nest() %>% 
  ungroup() %>%  
  mutate(
    model = furrr::future_map(data, 
      ~ with(.x, 
        pb_sol_wham(ph = pH_in, dic = 5, phase = "Hydcerussite", Na = 10 / chemr::mass("Na"), mass_ha = nom_in)
      )
    )
  )

nom_grid %>% 
  unnest(c(data, model)) %>% 
  filter(near(pH_in, pH, .1)) %>% 
  mutate(
    total_pb = mol_Hydcerussite * 1e6 * 3 * chemr::mass("Pb"),
    nom_in = 1e3 * nom_in, 
    pb_ppb = log10(total_pb)
  ) %>% 
  ggplot(aes(nom_in, pH_in, fill = pb_ppb)) + 
  scale_fill_gradientn(
    colours = palette, 
    breaks = log10(c(100, 300, 1000, 3e3)),
    labels = function(breaks) 10 ^ breaks
  ) +
  geom_raster() + 
  geom_contour(aes(z = pb_ppb), col = "white", linetype = 2) +
  labs(
    x = expression("Humic acid (mg C L"^-1*")"),
    y = "pH",
    fill = expression("[Pb] (µg L"^-1*")")
  )

```


