---
title: "Non-parametric matched pair testing with left-censored data"
description: |
  Comparing two groups of measurements when some values are below a detection limit.
author:
  - name: Ben Trueman
    url: {}
date: 07-05-2021
bibliography: references.bib
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

There are few options in R for comparing matched pairs in two groups with left-censored data. And while `NADA2::cen_signedranktest()` is an excellent tool, I wrote the following as another. It implements the Paired Prentice-Wilcoxon test, as described in @helsel_statistics_2012. I should also acknowledge the USGS' orphaned package smwrQW [@smwrQW] for its version of the same test.

```{r ppw-test, echo=TRUE}

library("tidyverse")
library("survival")

ppw_test <- function(
  x, # numeric
  y, # numeric
  x_cen, # logical 
  y_cen, # logical 
  alternative = "two.sided", # either "two.sided", "greater" (x > y), or "less" (x < y)
  flip = TRUE
) {
  
  if(length(x) != length(y))
    stop("Lengths of x and y must be the same for paired data.")
  
  if(any(c(x, y) < 0))
    stop("Negative values in x or y.")
  
  valid_alternative <- pmatch(alternative, c("two.sided", "greater", "less"))
  
  if(is.na(valid_alternative))
    stop('Invalid choice for alternative, must match "two.sided", "greater", or "less"')

  test_input <- tibble(x_val = x, y_val = y, x_cen, y_cen) %>% 
    na.omit() %>% 
    rowid_to_column() %>% 
    pivot_longer(
      cols = -rowid,
      names_to = c("group", ".value"),
      names_pattern = "(.+)_(.+)"
    ) %>% 
    mutate(
      cen = as.numeric(!cen), # 0 is censored, 1 is observed
      # flip data so that smallest observation becomes longest "survival time":
      # N.B., this rounds the flipped data to 6 decimal places
      flipped = if(flip) {max(val) + 1 - val} else val,
      flipped = round(flipped, 6)
    ) %>% 
    left_join(
      # estimate survival function:
      survival::survfit(survival::Surv(flipped, cen) ~ 1, data = .) %>% 
        broom::tidy() %>% 
        mutate(time = round(time, 6)),
      by = c("flipped" = "time")
    ) %>% 
    mutate(score = if_else(cen == 1, 1 - 2 * estimate, 1 - estimate)) %>% 
    pivot_wider(id_cols = rowid, names_from = group, values_from = score) %>% 
    mutate(d = x - y) %>% 
    summarize(
      z_ppw = sum(d) / sqrt(sum(d ^ 2)),
      p_val = if(alternative == "two.sided") {2 * pnorm(abs(z_ppw), lower.tail = FALSE)} else
        if(alternative == "greater") {pnorm(-z_ppw, lower.tail = FALSE)} else
          if(alternative == "less") {pnorm(z_ppw, lower.tail = FALSE)} else
            "alternative hypothesis is invalid"
    )
  
  list("statistic" = test_input$z_ppw, "p.value" = test_input$p_val)
  
}


```

The `ppw_test()` function works like this:

```{r basic-use}

withr::with_seed(450, { # generate two random normal variables, with left-censoring:
  tibble( 
    x = rnorm(10, 3, 1),
    y = rnorm(10, 3, 1),
    x_cen = x < 2,
    y_cen = y < 2
  )
}) %>% 
  with(ppw_test(x, y, x_cen, y_cen))
```
It also does one-sided tests:

```{r one-sided}

withr::with_seed(23, {
  tibble( 
    x = rnorm(10, 10, 1),
    y = rnorm(10, 5, 1),
    x_cen = x < 10,
    y_cen = y < 5
  )
}) %>% 
  with(ppw_test(x, y, x_cen, y_cen, alternative = "greater")) 

```

The following code tests that `ppw_test()` gives the expected result when applied to the data in Table 9.7 of @helsel_statistics_2012. First, here are the data:

```{r, echo=FALSE}
theme_set(theme_classic())
```

```{r helsel-data}
helsel <- read_csv("helsel_table_9.7 copy.csv")
```

```{r helsel-plot, fig.height=3, echo=FALSE}

helsel %>% 
  transmute(
    june_num = str_remove(june, "<") %>% 
      as.numeric(),
    june_cen = str_detect(june, "<"),
    sept_num = str_remove(september, "<") %>% 
      as.numeric(),
    sept_cen = str_detect(september, "<"),
  ) %>% 
  pivot_longer(everything(), 
    names_to = c("month", ".value"),
    names_pattern = "(.+)_(.+)"
  ) %>% 
  ggplot(aes(
    x = num, 
    y = str_to_sentence(month), 
    col = if_else(cen, "Nondetect", "Quantified")
  )) + 
  geom_jitter(height = .1, width = 0, size = 4, shape = 16, alpha = .5) + 
  scale_x_log10() + 
  scale_colour_manual(values = wesanderson::wes_palette("Zissou1")[c(1,5)]) +
  labs(
    x = expression("Atrazine (µg L"^-1*")"),
    y = NULL,
    col = NULL
  )

```

And here are the test results. These are close to the values in @helsel_statistics_2012, but not exactly the same. I suspect there are a few typos in the table, which may have something to do with it. For example, while most of the scores calculated by `ppw_test()` are consistent with those reported in the table, line 11, column 3 list a score of 0.55 for the second-highest value in column 1, while `ppw_test()` calculates a score of -0.54, much closer to the value of -0.67 corresponding to the largest value in column 1. There is a similar problem on line 12.

```{r helsel-test}

helsel %>% 
  mutate(
    june_cen = str_detect(june, "<"),
    sept_cen = str_detect(september, "<"),
  ) %>% 
  mutate_if(is.character, ~ as.numeric(str_remove(.x, "<"))) %>% 
  with(ppw_test(june, september, june_cen, sept_cen, "less"))

```

<!-- The following code tests that `ppw_test()` gives the expected result on the data of @obrien_paired_1987. -->

```{r o-brien-test, eval=FALSE}

tibble::tribble(
  # from o'brien and fleming, 1987:
  ~patient, ~close_match, ~poor_match,
  1, "37", 29, 
  2, "19", 13, 
  3, ">57", 15, 
  4, "93", 26, 
  5, "16", 11, 
  6, "22", 17, 
  7, "20", 26, 
  8, "18", 21, 
  9, "63", 43, 
  10, "29", 15,
  11, ">60", 40
) %>% 
  mutate(
    cen_close = str_detect(close_match, ">"),
    cen_poor = FALSE,
    close_match = str_remove(close_match, ">") %>% 
      as.numeric()
  ) %>% 
  with(ppw_test(close_match, poor_match, cen_close, cen_poor, alternative = "less", flip = FALSE))

```

```{r inspect-helsel, eval=FALSE}

test <- helsel %>% 
  transmute(
    june_num = str_remove(june, "<") %>% 
      as.numeric(),
    june_cen = str_detect(june, "<"),
    sept_num = str_remove(september, "<") %>% 
      as.numeric(),
    sept_cen = str_detect(september, "<"),
  )

x <- test$june_num
y <- test$sept_num
x_cen <- test$june_cen
y_cen <- test$sept_cen

alternative <- "two.sided"
flip <- TRUE

test_input <- tibble(x_val = x, y_val = y, x_cen, y_cen) %>% 
    na.omit() %>% 
    rowid_to_column() %>% 
    pivot_longer(
      cols = -rowid,
      names_to = c("group", ".value"),
      names_pattern = "(.+)_(.+)"
    ) %>% 
    mutate(
      cen = as.numeric(!cen), # 0 is censored, 1 is observed
      # flip data so that smallest observation becomes longest "survival time":
      # N.B., this rounds the flipped data to 6 decimal places
      flipped = if(flip) {max(val) + 1 - val} else val,
      flipped = round(flipped, 6)
    ) %>% 
    left_join(
      # estimate survival function:
      survival::survfit(survival::Surv(flipped, cen) ~ 1, data = .) %>% 
        broom::tidy() %>% 
        mutate(time = round(time, 6)),
      by = c("flipped" = "time")
    ) %>% 
    mutate(score = if_else(cen == 1, 1 - 2 * estimate, 1 - estimate)) %>% 
    pivot_wider(id_cols = rowid, names_from = group, values_from = score) %>% 
    mutate(d = x - y)

tibble(
  jun_helsel = helsel$june_score_s1,
  jun_ben = test_input$x,
  sep_helsel = helsel$june_score_s1,
  sep_ben = test_input$x,
  d_helsel = helsel$diff_s1_s2,
  d_ben = test_input$d
) %>% 
  rowid_to_column() %>% 
  pivot_longer(-rowid,
    names_to = c("set", ".value"),
    names_pattern = "(.+)_(.+)"
  ) %>% 
  mutate(ben = round(ben, 2)) %>% 
  filter(!near(helsel, ben, .1))

```

