---
title: "Peak detection for qualitative XRD analysis in R"
description: |
  Improved visualization of X-ray diffraction data.
author:
  - name: Ben Trueman
    url: 
date: 06-29-2021
bibliography: references.bib
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load, echo=FALSE}
library("tidyverse")
library("fffprocessr") # for peak detection
theme_set(
  theme_classic() + 
    theme(
      axis.line.y = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
    )
)

unit_scale <- . %>% # scale to [0,1]
  mutate(
    intensity = intensity - min(intensity, na.rm = TRUE),
    intensity = intensity / max(intensity, na.rm = TRUE)
  )

data <- read_csv("xrd_data_digestion.csv") %>% 
  filter(sample == "lead_carbonate") %>% 
  unit_scale() 
  

stds <- read_csv("xrd_standards.csv") %>% 
  filter(
    phase %in% c("Cerussite", "Hydrocerussite"),
    two_theta >= min(data$two_theta),
    two_theta <= max(data$two_theta)
  ) %>% 
  unit_scale() 
```

A important aspect of understanding inorganic contaminant mobility in drinking water systems is identifying the solid phases that exist on the surfaces of pipes. This is often done using X-ray diffraction (XRD), and I am often preparing figures to communicate XRD data. This post provides an example workflow for displaying XRD data and labelling the peaks so that readers can easily attribute them. It makes use of the peak detection function `peak_maxima()` in my field-flow fractionation data analysis package `fffprocessr` [@fffprocessr].

For this example, we'll use an XRD pattern representing a mixed lead carbonate phase containing cerussite (PbCO~3~) and hydrocerussite (Pb~3~(CO~3~)~2~(OH)~2~). Here are the data:

```{r plot-funs, fig.height=2.5}

label_axes <- function(...) {
  labs(
    x = expression("2"*theta~"(Cu K"*alpha*")"),
    y = NULL,
    ...
  )
}

custom_colour_scale <- function(n = 3, ...) {
  scale_colour_manual(values = wesanderson::wes_palette("Zissou1", n)[c(1, n)])
}

data %>%
  ggplot(aes(two_theta, intensity)) +
  geom_line() +
  label_axes()
```

First we need a quick and dirty method of determining the order in which phases should appear in the plot---interpretation is easiest when the standard that matches the data best appears closest to it. 

```{r correlations}

xrd_corr <- function(run, standard) {
  list( 
      sample = run, 
      std = standard
    ) %>% 
    bind_rows(.id = "type") %>% 
    pivot_wider(id_cols = two_theta, names_from = type, values_from = intensity) %>% 
    group_by(two_theta = round(two_theta)) %>% 
    summarize(sample = median(sample, na.rm = TRUE), std = median(std, na.rm = TRUE)) %>% 
    ungroup() %>% 
    with(cor(sample, std, use = "complete", method = "pearson"))
}

importance <- stds %>% 
  distinct(phase) %>% 
  pull(phase) %>% 
  set_names() %>% 
  map_dfc(
    ~ xrd_corr(
        run = data,
        standard = filter(stds, phase == .x) %>% distinct()
    )
  ) %>% 
  pivot_longer(everything(), names_to = "phase", values_to = "r") %>% 
  arrange(desc(r))

```

Here are the data with the standard patterns for cerussite and hydrrocerussite. Cerussite is a better match to the data and so it is plotted closest to the data.

```{r data-w-stds, fig.height=3}

ordered_stds <- stds %>% 
  mutate(phase_f = factor(phase) %>% fct_relevel(importance$phase) %>% as.numeric())

data %>% 
  ggplot(aes(two_theta, intensity)) + 
  geom_line() +
  geom_segment(
    data = ordered_stds,
    aes(
      x = two_theta, xend = two_theta, 
      y = .25 * (0 - phase_f), yend = .25 * (intensity - phase_f), 
      col = phase
    )
  ) + 
  scale_y_continuous(breaks = seq(0, 1, .25)) +
  label_axes(col = NULL) + 
  custom_colour_scale(4)

```
We detect peaks in the pattern using `fffprocessr::peak_maxima()`.

```{r detect}

peaks_detected <- data %>% 
  peak_maxima(
    peaks = 23, n = 1, method = "sigma", 
     x_var = "two_theta", y_var = "intensity",
    group_vars = NULL
  )
```

Then, we need a function to assign the detected peaks to the appropriate standard.

```{r assign}

assign_peaks <- function(sample, standard, tol = 1, phases) { 
  sample %>% 
    mutate(two_theta_rnd = plyr::round_any(two_theta, tol)) %>% 
    right_join(
      standard %>% 
        filter(phase %in% phases) %>% 
        mutate(two_theta_rnd = plyr::round_any(two_theta, tol)) %>% 
        select(two_theta, two_theta_rnd, phase), 
      by = "two_theta_rnd", suffix = c("", "_std")
    ) %>% 
    group_by(two_theta = round(two_theta_std, 1), phase) %>%
    summarize(intensity = max(intensity)) %>%
    ungroup()
}

peaks_idd <- bind_rows(
  assign_peaks(peaks_detected, stds, phases = importance$phase[1], tol = .5),
  assign_peaks(peaks_detected, stds, phases = importance$phase[2], tol = .5)
)

```

Finally, we add the identified peaks to the plot:

```{r data-w-peaks, fig.height=3.5}

data %>% 
  ggplot(aes(two_theta, intensity)) + 
  geom_line() +
  geom_segment(
    data = ordered_stds,
    aes(
      x = two_theta, xend = two_theta, 
      y = .25 * (0 - phase_f), yend = .25 * (intensity - phase_f), 
      col = phase
    )
  ) + 
  geom_point(
    data = peaks_idd,
    aes(col = phase)
  ) + 
  scale_y_continuous(breaks = seq(0, 1, .25)) +
  label_axes(col = NULL) + 
  custom_colour_scale(4)

```


