---
title: "Diagnostics for censored autoregressive models fitted with brms"
description: |
  Posterior predictive checks and simulated residuals.
author:
  - name: Ben Trueman
    url: {}
date: 10-08-2021
output:
  distill::distill_article:
    self_contained: false
bibliography: references.bib
---

```{r setup, include=FALSE}
here::i_am("2021-10-08-diagnostics-for-censored-autoregressive-models-fit-with-brms/diagnostics-for-censored-autoregressive-models-fit-with-brms.Rmd")

knitr::opts_chunk$set(echo = FALSE)
library("tidyverse")
library("NADA")
library("brms")

theme_set(
  theme_bw(14) + 
    theme(
      legend.position = "bottom",
      strip.background = element_blank()
    )
)
pal <- wesanderson::wes_palette("Zissou1")
```

```{r load, message=FALSE}
ldat <- read_csv(here::here("2021-09-08-a-brownification-pause/UPPER-SILVER-LAKE.csv"))
```

In a previous post ([A brownification pause](a-brownification-pause.html)), I made an attempt at tackling a common problem in environmental science: analyzing autocorrelated time series with left-censored values (i.e., nondetects). As I've learned, one powerful tool for this type of problem is `brms` [@brms; @brmsadv], an R package for fitting Bayesian regression models via Stan [@rstan].

There are, however, relatively few tools that I'm aware of for posterior predictive checks of censored regression models. The R function `bayesplot::ppc_km_overlay()` is one, but it is only suitable for right-censored data, which are less common in environmental time series.

Here I use a similar approach to generate a posterior predictive check for a left-censored model. I use the R function `NADA::cenfit()` [@nada] to estimate the empirical cumulative distribution function (ECDF) of the series and the posterior draws from the model. The function works by "flipping" the input---subtracting all values from a constant larger than any value---and estimating the ECDF according to the Kaplan-Meier method (for right-censored data).

```{r model-in}

ldat_clean <- ldat %>%
  arrange(date) %>% 
  transmute(
    date, 
    yr = lubridate::year(date),
    sem = lubridate::semester(date),
    value,
    ci = if_else(flag == "<", -1, 0),
    ci = replace_na(ci, 0)
  )

model_in <- crossing(
  yr = seq(1983, max(ldat_clean$yr), by = 1),
  sem = 1:2
) %>% 
  left_join(ldat_clean, by = c("yr", "sem")) %>%
  group_by(yr, sem) %>% 
  summarize(
    date = median(date),
    med_ci = median(ci),
    value = median(value)
  ) %>% 
  ungroup() %>% 
  transmute(
    date,
    numeric_date = yr + sem / 2,
    numeric_date = numeric_date - min(numeric_date) + 1,
    sem = factor(sem),
    value,
    ci = sign(med_ci),
    # replace missing values in the censoring indicator with 0 (uncensored)
    ci = replace_na(ci, 0),
    lcl = if_else(ci == -1, value, 5)
  )

```

```{r model-brms, echo=FALSE, results="hide"}

model_brms <- brm(
  bf(log(value) | cens(ci) + mi() ~
       s(numeric_date) + sem + ar(time = numeric_date, p = 1)),
  data = model_in,
  control = list(adapt_delta = .999),
  seed = 3152,
  file = here::here("2021-09-08-a-brownification-pause/model_brms"),
  cores = 4
)

model_brms_noar <- brm(
  bf(log(value) | cens(ci) + mi() ~ s(numeric_date) + sem),
  data = model_in,
  control = list(adapt_delta = .999),
  seed = 3152,
  file = here::here(
    "2021-10-08-diagnostics-for-censored-autoregressive-models-fit-with-brms/model_brms_noar"
  ),
  cores = 4
)

```

The following generates ECDFs of the data and posterior predictions according to this method:

```{r ecdf, echo=TRUE}

pp_ecdf <- function(model, newdata, yval, log_t = TRUE) {
  
  ecdf_data_in <- newdata %>% 
  # convert censoring indicator to logical:
  mutate(ci = ci == -1) %>% 
  filter(!is.na({{yval}}))
  
  ecdf_data <- NADA::cenfit(
    obs = if(log_t) {
      log(pull(ecdf_data_in, {{yval}}))
    } else {
      pull(ecdf_data_in, {{yval}})
    },
    censored = ecdf_data_in$ci
  )

ecdf_pp <- tidybayes::add_predicted_draws(
    newdata, 
    model,
    ndraws = 200
  ) %>% 
  ungroup() %>% 
  group_by(.draw) %>% 
  nest() %>% 
  ungroup() %>% 
  mutate(
    cenfit = map(data, 
      ~ with(.x, 
        NADA::cenfit(
          obs = .prediction, 
          censored = rep(FALSE, length(.prediction))
        )
      )
    ),
    cenfit_summ = map(cenfit, summary)
  ) %>% 
  unnest(cenfit_summ) %>% 
  select(where(~ !is.list(.x))) 

  bind_rows(
    "Posterior draws" = ecdf_pp,
    "Observations" = summary(ecdf_data),
    .id = "type"
  )
}

```

Here I've superimposed the ECDF of the time series on the ECDFs estimated using 200 draws from the posterior distribution of the `brms::brm()` model. From this plot, it appears that the posterior draws approximate the data reasonably well.

```{r ecdf-plot}

bind_rows(
  "AR(1)" = pp_ecdf(model_brms, model_in, value),
  .id = "model"
) %>% 
  ggplot(
    aes(obs, prob, col = type, size = type, group = .draw)
  ) +
  scale_color_manual(values = pal[c(3,1)]) +
  scale_size_manual(values = c(.7, .2)) +
  geom_line() + 
  guides(size = "none") +
  labs(x = "Observation", y ="ECDF", col = NULL)

```

Another difficulty in evaluating models fitted to censored time series is residuals analysis. Here I adopt the approach of @wang_quasilikelihood_2018, generating simulated residuals by substituting censored and missing values of the time series with a draw from the posterior distribution of the fitted model and refitting the model on the augmented data. I then generated residual draws from the updated model.

The function below does the simulation:

```{r resids-simulate-fun, echo=TRUE}

simulate_residuals <- function(
  model, 
  newdata, 
  yval, 
  file, 
  seed = NULL,
  ...
) {
  
  data_aug <- tidybayes::add_predicted_draws(
    newdata,
    model,
    ndraws = 1,
    seed = seed
  ) %>% 
    ungroup() %>% 
    mutate(
      value = if_else(
        is.na({{yval}}) | ci == -1,
        .prediction,
        {{yval}}
      ),
      ci = 0 # no censoring
    ) %>%
    select(-starts_with("."))
  
  model_update <- update(
    model, 
    newdata = data_aug,
    file = file,
    cores = 4,
    seed = seed,
    ...
  )
  
  model_resids <- tidybayes::add_residual_draws(
    object = model_update,
    newdata = data_aug,
    method = "posterior_predict"
  )
  
  list(
    model = model_update, 
    residuals = model_resids, 
    data = data_aug
  )
  
}

```

```{r resids-simulate}

brm_seed <- 3125

resids <- simulate_residuals(
  model_brms, model_in, value,
  file = here::here("2021-10-08-diagnostics-for-censored-autoregressive-models-fit-with-brms/model_brms_resid"),
  seed = brm_seed
)

resids_noar <- simulate_residuals(
  model_brms_noar, model_in, value,
  file = here::here("2021-10-08-diagnostics-for-censored-autoregressive-models-fit-with-brms/model_brms_resid_noar"),
  control = list(adapt_delta = .999),
  seed = brm_seed
)
```

Here is the density of the lag one autocorrelation, estimated using residual draws from Bayesian GAMs fitted with and without a first-order autoregressive (AR(1)) term. Weak first-order autocorrelation is apparent in the residuals of the GAM without an AR(1) term and absent from those of the GAM with one.

```{r resids-plot}

bind_rows(
  "AR(1)" = resids$residuals,
  "Independent\nerrors" = resids_noar$residuals,
  .id = "model"
) %>% 
  ungroup() %>% 
  group_by(model, .draw) %>% 
  mutate(lag1 = dplyr::lag(.residual, 1)) %>% 
  # pivot_longer() allows for additional lags to be included above
  pivot_longer(
    starts_with("lag"), 
    names_to = "lag", 
    values_to = ".residual_lagged"
  ) %>% 
  group_by(lag, .add = TRUE) %>% 
  summarize(
    cor = cor(.residual_lagged, .residual, use = "complete")
  ) %>% 
  ungroup() %>% 
  ggplot(aes(cor, model)) + 
  # facet_wrap(vars(lag)) + # use this for multiple lags
  ggdist::stat_halfeye(
    aes(fill = stat(x < 0)), 
    slab_alpha = .5
  ) + 
  scale_fill_manual(values = pal[c(3, 1)]) +
  labs(x = "Lag 1 autocorrelation", y = NULL)
  
```

For a bit more verification, I fitted a similar GAM to a simulated dataset, generated by adding an AR(1) series to a nonlinear trend, as follows:

```{r data-simulate, echo=TRUE}
lcl <- 10 # lower censoring limit

simdat <- withr::with_seed(101, {
  tibble(
  x = 1:200,
  y_t = 1e-3 * x + 1e-4 * x ^ 2,
  e = arima.sim(list(ar = .5), length(y_t)) %>% 
    as.numeric(),
  y = y_t + e + 10,
  y_star = pmax(y, lcl),
  ci = if_else(y < lcl, -1, 0)
)
})
```

Again, I fitted the GAM---and the equivalent model without an AR(1) 
term---using `brms`:

```{r filepath}
file_path <- "2021-10-08-diagnostics-for-censored-autoregressive-models-fit-with-brms"
```

```{r model, echo=TRUE}

model_simdat <- brm(
  bf(y_star | cens(ci) ~ s(x) + ar(time = x, p = 1)),
  data = simdat,
  seed = 124,
  # save the model:
  file = here::here(paste(file_path, "model_simdat", sep = "/")),
  cores = 4,
  control = list(adapt_delta = .99)
)

```

Here are the simulated data and the fitted model:

```{r data-plot}

fitted_vals <- fitted(model_simdat) %>% 
  as_tibble()

simdat %>% 
  bind_cols(fitted_vals) %>% 
  pivot_longer(c(y, y_star, Estimate)) %>% 
  mutate(name = fct_relevel(name, "Estimate", after = Inf)) %>% 
  ggplot(aes(x)) +
  scale_color_manual(
    values = c("grey", "black", pal[1]),
    labels = c(
      Estimate = "Model", 
      y = "Uncensored", 
      y_star = "Censored"
    )
  ) +
  geom_line(aes(y = value, col = name)) + 
  geom_ribbon(
    data = function(x) x %>% 
      filter(name == "Estimate"),
    aes(ymin = Q2.5, ymax = Q97.5),
    alpha = .5, col = NA, fill = pal[1]
  ) +
  labs(x = NULL, y = NULL, col = NULL)

```

```{r model-no-ar}

model_simdat_noar <- brm(
  bf(y_star | cens(ci) ~ s(x)),
  data = simdat,
  file = here::here(paste(file_path, "model_simdat_noar", sep = "/")),
  cores = 4,
  control = list(adapt_delta = .99),
  seed = 124
)

```

And here is the ECDF overlay described above:

```{r ecdf-sim, preview=TRUE}

pp_ecdf(model_simdat, simdat, y_star, log_t = FALSE) %>% 
  ggplot(
    aes(obs, prob, col = type, size = type, group = .draw)
  ) +
  scale_color_manual(values = pal[c(3,1)]) +
  scale_size_manual(values = c(.7, .2)) +
  geom_line() + 
  guides(size = "none") +
  labs(x = "Observation", y ="ECDF", col = NULL)

```

```{r resid-sim}

resids_simdat <- simulate_residuals(
  model_simdat, simdat, y_star,
  file = here::here("2021-10-08-diagnostics-for-censored-autoregressive-models-fit-with-brms/model_simdat_resid"),
  seed = brm_seed
)

resids_simdat_noar <- simulate_residuals(
  model_simdat_noar, simdat, y_star,
  file = here::here("2021-10-08-diagnostics-for-censored-autoregressive-models-fit-with-brms/model_simdat_resid_noar"),
  seed = brm_seed
)

```

The first-order autocorrelation estimate from the simulated residuals suggests that the model is accounting for autocorrelation in the residuals, as expected:

```{r resdid-sim-plot}

bind_rows(
  "AR(1)" = resids_simdat$residuals,
  "Independent\nerrors" = resids_simdat_noar$residuals,
  .id = "model"
) %>% 
  ungroup() %>% 
  group_by(model, .draw) %>% 
  mutate(
    lag1 = dplyr::lag(.residual, 1),
    # lag2 = dplyr::lag(.residual, 2),
    # lag3 = dplyr::lag(.residual, 3),
    # lag4 = dplyr::lag(.residual, 4)
  ) %>% 
  pivot_longer(
    starts_with("lag"), 
    names_to = "lag", 
    values_to = ".residual_lagged"
  ) %>% 
  group_by(lag, .add = TRUE) %>% 
  summarize(
    cor = cor(.residual_lagged, .residual, use = "complete")
  ) %>% 
  ungroup() %>% 
  ggplot(aes(cor, model)) + 
  # facet_wrap(vars(lag)) +
  ggdist::stat_halfeye(
    aes(fill = stat(x < 0)), 
    slab_alpha = .5
  ) + 
  scale_fill_manual(values = pal[c(3, 1)]) +
  labs(x = "Lag 1 autocorrelation", y = NULL)

```

And the estimate of the AR(1) term is `r round(summary(model_simdat)$cor_pars$Estimate, 2)`, which is similar to true value of 0.5.

