---
title: "Building a continuous-time autoregressive model in brms"
description: |
  A simple modification of brms-generated Stan code to fit first-order autoregressive models to irregularly-spaced time series.
author:
  - name: Ben Trueman
    url: {}
date: 02-01-2022
output:
  distill::distill_article:
    self_contained: false
bibliography: references.bib
---


```{r setup, include=FALSE}
here::i_am("building-a-continuous-time-autoregressive-model-in-brms-and-stan.Rmd")

knitr::opts_chunk$set(echo = TRUE)

library("dplyr")
library("stringr")
library("tidyr")
library("ggplot2")
library("brms")
library("rstan")

options(mc.cores = parallel::detectCores())

theme_set(
  theme_bw(14) + 
    theme(
      legend.position = "bottom",
      strip.background = element_blank()
    )
)
pal <- wesanderson::wes_palette("Zissou1")
```

EDIT: This post has been updated to reflect the changes to prior specification implemented in `brms 2.17.0`.

In recent posts, I've written about using `brms` to fit fully Bayesian autoregressive models to left-censored time series data. And while this approach is powerful, it is not easy to use for irregularly spaced series.

It turns out, though, that there is a straightforward generalization of the first-order autoregressive---AR(1)---model, called the continuous-time AR(1), or CAR(1). Whereas the AR(1) takes the form 

$$x_t = \phi x_{t-1} + \epsilon_t$$

(where $x$ is the time series, $t$ is time, $\phi$ defines the autocorrelation structure, and $\epsilon_t$ is an independent error term), the CAR(1) model has the following autocorrelation structure:

$$h(s, \phi) = \phi^s, s\geq0, \phi\geq0 $$
where $s$ is a real number representing the time difference between successive observations. This model is implemented in R by the function `nlme::corCAR1()` [@nlme], and there is an open issue on GitHub discussing implementation in [`brms`](https://github.com/paul-buerkner/brms/issues/741). 

Not being able to wait for a future version of `brms` with CAR(1) as an option, I modified the `brms` generated Stan code to fit a CAR(1) model, as follows.

First, let's simulate a couple of irregularly-spaced AR(1) processes:

```{r simulate}

stan_seed <- 1256

phi <- .75
p_ret <- .6 # proportion retained

withr::with_seed(stan_seed, {
  data <- tibble(
    x = 1:100,
    y1 = arima.sim(list(ar = phi), length(x)),
    y2 = arima.sim(list(ar = phi), length(x))
  ) %>% 
    pivot_longer(starts_with("y"), names_to = "g", values_to = "y")
  
  subset <- data %>% 
    slice_sample(prop = p_ret) %>% 
    arrange(g, x) %>% 
    group_by(g) %>% 
    mutate(
      x_lag = lag(x),
      d_x = replace_na(x - x_lag, 0) # spacing of observations
    ) %>% 
    ungroup()
})

```

Then, use `brms` to generate Stan code and the accompanying data as a list:

```{r generate-code}

priors <- prior(normal(.5, .25), class = ar, lb = 0, ub = 1)
formula <- bf(y ~ ar(time = x, gr = g))

sdata <- brms::make_standata(formula, prior = priors, data = subset)
sdata$s <- subset$d_x # CAR(1) exponent 

scode <- brms::make_stancode(formula, prior = priors, data = subset)

```

Next, modify the Stan code to fit a CAR(1) model... 

```{r modify}

scode_car1 <- scode %>% 
  # add time difference variable s:
  str_replace(
    "response variable\\\n", 
    "response variable\n  vector[N] s;  // CAR(1) exponent\n"
  ) %>% 
  # set lower bound of zero on ar param:
  str_replace(
    "vector\\[Kar\\] ar;", 
    "vector<lower=0>[Kar] ar;"
  ) %>% 
  # convert AR process to CAR1:
  str_replace(
    "mu\\[n\\] \\+= Err\\[n, 1:Kar\\] \\* ar;", 
    "mu[n] += Err[n, 1] * pow(ar[1], s[n]); // CAR(1)"
  ) 

class(scode_car1) <- "brmsmodel"

```

... and pass the data list and the modified Stan code to `rstan::stan()` to fit the model.

```{r fit, results="hide", eval=FALSE}

stanfit <- rstan::stan(
  model_code = scode_car1,
  data = sdata,
  sample_file = "carmodel", # output in csv format
  seed = stan_seed
)

```

```{r load-fit, echo=FALSE}

csvfiles <- list.files(
  pattern = "carmodel_\\d\\.csv",
  full.names = TRUE
)

stanfit <- rstan::read_stan_csv(csvfiles)

```

Once the chains have finished running, feed the Stan model back into `brms`, and have a look at the output of `summary()`. There are no obvious problems with convergence, given the lack of warnings from Stan and the $\widehat{r}$ values being very close to 1.

```{r as-brms}

fit <- brm(formula, data = subset, empty = TRUE)
fit$fit <- stanfit
fit <- rename_pars(fit)
summary(fit)

```

Since the `brmsfit` object doesn't contain the CAR(1) formula, a few extra steps are needed to generate predictions. First, we generate draws from the model without the autocorrelation structure, then we apply a CAR(1) filter to the data, and then we summarize the filtered draws using `median_qi()`. Plot the medians along with the data, adding the 0.025 and 0.975 quantiles as a ribbon:

```{r preds}

# extract AR(1) draws:
phi <- as_draws_df(fit, "ar[1]") %>% 
  as_tibble()

# generate draws from the model without the autocorrelation structure:
pred_car1 <- tidybayes::add_epred_draws(subset, fit, incl_autocor = FALSE) %>% 
  ungroup() %>% 
  select(-c(.chain, .iteration)) %>% 
  arrange(.draw, g, x) %>% 
  left_join(phi, by = ".draw") %>% 
  # add the CAR(1) structure:
  group_by(.draw, g) %>% 
  mutate(
    r_lag = replace_na(lag(y - .epred), 0),
    .epred = .epred + r_lag * `ar[1]` ^ d_x
  ) %>% 
  ungroup() %>% 
  # summarize:
  select(-c(r_lag, `ar[1]`)) %>% 
  group_by(across(matches(paste(names(subset), collapse = "|")))) %>% 
  ggdist::median_qi() %>% 
  ungroup()
```


```{r plot, echo=FALSE}

pred_car1 %>%
  mutate(g = forcats::fct_recode(g, "Series 1" = "y1", "Series 2" = "y2")) %>% 
  ggplot(aes(x)) + 
  facet_wrap(vars(g), ncol = 1) + 
  geom_line(aes(y = y, col = "Data")) + 
  geom_point(aes(y = y, col = "Data")) + 
  geom_ribbon(
    aes(ymin = .lower, ymax = .upper, fill = "Model"),
    alpha = .4, show.legend = FALSE
  ) +
  geom_line(aes(y = .epred, col = "Model")) + 
  scale_color_manual(values = pal[c(2,5)]) + 
  labs(x = NULL, y = NULL, col = NULL)

```

And that's it! The model recovers the parameters used to generate the simulated data well in this case---the mean of the posterior of $\phi$ is `r round(posterior::summarise_draws(phi)$median, 2)`---which is a good sign that we're on the right track.

